A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation\tnoteref{ornl-cr}

### 2013-05-16 doc/jcomp2012/jcomp2012.tex
Reviewer edits.

@@

@@

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three-dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the one-temperature (1T) thermal radiation
  diffusion equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

@@

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$:
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events that initiates in state $i$:
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk. In
this method, the weight change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are normalized
over a column, as opposed to the forward method in which the probabilities are
normalized over a row.  This is equivalent to forming the Neumann series in
reverse order.  Correspondingly, the estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\ &=
    \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in state $i_0$.
The Kronecker delta implies that tallies are only made in the state where the
random walk currently resides.  This is the common approach in standard Monte
Carlo transport simulations. Thus, unlike the direct method, we are no longer
required to start a history in each state for each random walk permutation
because tallies are made in the state in which the random walk currently
resides. Instead, sampling the source is sufficient, and one random walk per
permutation is required.

@@

We propose a modification of this scheme that uses Monte Carlo as a
synthetic acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
from Eq.~(\ref{eq:point_iteration}):
\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}
where
\begin{equation}
  \delta \vx^{l} = \vx - \vx^{l}
\end{equation}
is defined as the error at iteration $l$. We then subtract $(\vI-\vA)
\delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear
system to solve for the error:
\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\ &=
    \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme that applies the Monte Carlo method to estimate
the solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:MCSA}
\end{subequations}
As in the sequential method, the hat on $\vA$ in
Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the Monte Carlo solution
only approximately inverts this operator.  Thus, we have defined a scheme in
which the initial estimate of $\vx$ in each iteration is updated using a
single fixed-point iteration\footnote{A subspace method could be used instead
  of a stationary method here.}.  The residual is calculated and used as a
source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
error is then used to calculate an updated iterate of the solution vector,
$\vx^{l+1}$.  The entire sequence is iterated to convergence based on the
following stopping criterion~\cite{kelley_1995}:
\begin{equation}
  \|\vr\|_{2} < \epsilon\cdot\|\vb\|_2\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

Before completing the derivation of discrete equations required to solve
Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary conditions.  For
this work, the Marshak Boundary condition suffices:
\begin{equation}
  f_b^\pm = \frac{1}{4}\phi -
  \frac{1}{2}\ve{F}\cdot\hat{\ve{n}}\:,\quad \vr\in \partial\Gamma\:,
\end{equation}
where $f_b$ is the incoming partial current on the problem boundary,
$\partial\Gamma$.  On any low boundary we can write the flux using a
first-order discrete form of Fick's Law:
\begin{equation}
  F_{1/2} = -\frac{2}{3\sigma_1}\frac{\phi_1 -
    \phi_{1/2}}{\Delta_1}\:.
  \label{eq:low-boundary-flux}
\end{equation}
As before, we enforce continuity of flux on the boundary using the
Marshak Boundary condition:
\begin{equation}
  -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1} = 2f_b^{+}
  -\frac{1}{2}\phi_{1/2}\:.
\end{equation}
Solving for $\phi_{1/2}$ and applying to Eq.~(\ref{eq:low-boundary-flux})
yields an expression for the low-side boundary flux:
\begin{equation}
  F_{1/2} = \frac{2}{3\sigma_1\Delta_1 + 4}(4f_b^{+}-\phi_1)\:.
\end{equation}
Following an identical procedure on high-side problem boundaries gives
\begin{equation}
  F_{N_l+1/2} = \frac{2}{3\sigma_{N_l}\Delta_{N_l} +
    4}(\phi_{N_l}-4f_b^{-})\:,
\end{equation}
where $l\in\{i,j,k\}$ is defined over the range $[1,N_l]$.

@@

The adjoint Monte Carlo method described in \S~\ref{sec:adjoint-method} is
used to estimate
\begin{equation}
  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
\end{equation}
As noted above, the residual acts as the source for this simulation, and the
operator $\ve{M}^{-1}\hat{\ve{D}}$ is implicitly defined in
Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
interpretations that can be applied to this system.  The first follows the
mathematical presentation given in \S~\ref{sec:monte-carlo-matrix}.  Namely,
we form probabilities from the iteration matrix and perform random walks to
generate unbiased estimates of the solution vector, $\delta\bphi^{l+1/2}$,
using the estimator in Eq.~(\ref{eq:adjoint-tally}).

@@

The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
Figure~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
collision results in a scatter to one of the six adjacent cells since
$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
probabilities for each scattering event from
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
the iteration matrix.  In this mesh cell, the probabilities for leakage out of
each face are derived from the equations from each adjacent cell.  Using
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
equations from each adjacent cell, here only noting the contribution from cell
7:
\begin{align*}
  \phi_1 &= \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}
  \frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\
  %%
  \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
  {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
  %%
  \phi_3 &= \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}
  \frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\
  %%
  \phi_4 &= \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}
  \frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
  %%
  \phi_5 &= \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}
  \frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
  %%
  \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:.\\
\end{align*}
A quick view of these equations reveals that the only non-zero entries in
column 7 of the iteration matrix are
\begin{equation*}
  \begin{pmatrix}
    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
  \end{pmatrix}^{T}
  =
  \begin{pmatrix}
    \frac{\sigm_{7\rightarrow
        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
    \\ \frac{\sigp_{7\rightarrow
        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
    \\ \frac{\sigm_{7\rightarrow
        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
    \\ \frac{\sigp_{7\rightarrow
        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
    \\ \frac{\sigm_{7\rightarrow
        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
    \\ \frac{\sigp_{7\rightarrow
        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
  \end{pmatrix}\:,
\end{equation*}
where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
cell 7 as
\begin{equation*}
  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
  \{1,\ldots,6\}\:.
\end{equation*}
The weight change for each scattering event is calculated using
Eq.~(\ref{eq:adjoint-weight}):
\begin{equation*}
  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
  \{1,\ldots,6\}\:,
\end{equation*}
where $m$ is the step-index in the particle history.  We also note that, for
the 3D Cartesian mesh considered in this work, the maximum number of non-zero
entries in columns of the iteration matrix is 6, generating a sparse system.
Cells along the problem boundary can have fewer than six entries.
\begin{figure}[ht!]
  \centerline{ \includegraphics[clip]{3D_mesh_segment.pdf}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

>>>>

  \author[ornl]{Steven P. Hamilton\fnref{fn1}} \ead{hamiltonsp@ornl.gov}

@@

In nearly all computational engineering and physics fields, linear and
non-linear solvers form the core components of modeling and simulation
applications. Recent focus on multiphysics coupling adds additional complexity
to common linear and non-linear systems as solution strategies change when
physical models are coupled. Furthermore, a desire for predictive simulations
to enhance the safety and performance of engineered systems creates a need for
extremely high fidelity computations to be performed for these coupled systems
as a means to capture effects not modeled by coarser methods. In order to
achieve this high fidelity, state-of-the-art computing facilities must be
leveraged in a way that is both efficient and considerate of hardware-related
issues. As scientific computing moves towards exascale facilities, with
machines of $O(1,000,000)$ cores already coming online, new algorithms to
solve these complex problems must be developed to leverage this new
hardware. Issues such as resiliency to node failures and scaling to large
numbers of heterogeneous computing elements (CPUs and GPUs) will be pertinent
to robust algorithms aimed at this new hardware.

@@

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three-dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the one-temperature (1T) thermal radiation
  diffusion equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}
While parallel scaling and resiliency on Petascale systems and beyond provides
strong motivation for pursuing this work, we do not directly address these
issues in this study.  These topics will be the subject of future
investigations.

@@

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
Using this series, we can write an iterative method that solves
Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$:
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
Here, $m$ is the random walk step index and the $h$ terms are the matrix
elements of $\vH$.  The terms in the Neumann series can be interpreted as a
series of transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk with $k$
events that initiates in state $i$:
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk. In
this method, the weight change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are normalized
over a column, as opposed to the forward method in which the probabilities are
normalized over a row.  This is equivalent to forming the Neumann series in
reverse order.  Correspondingly, the estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\ &=
    \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in state $i_0$.
The Kronecker delta implies that tallies are only made in the state where the
random walk currently resides.  Thus, the adjoint method is equivalent to the
common approach in standard Monte Carlo transport simulations. As opposed to
the direct method, the adjoint method does not require starting a history in
each state for each random walk permutation because tallies are made in the
state in which the random walk currently resides. Instead, sampling the source
is sufficient, and one random walk per permutation is required.

@@

We propose a modification of this scheme that uses Monte Carlo as a synthetic
acceleration \cite{dsa} for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
from Eq.~(\ref{eq:point_iteration}):
\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}
where
\begin{equation}
  \delta \vx^{l} = \vx - \vx^{l}
\end{equation}
is defined as the error at iteration $l$. We then subtract $(\vI-\vA)
\delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear
system to solve for the error:
\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\ &=
    \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
These steps can be written as an accelerated iteration scheme that applies the
Monte Carlo method to estimate the error in each iteration. The Monte Carlo
method is used to approximately invert $\vA$ in
Eq.~(\ref{eq:synthetic-residual-solve}). This updated scheme, the Monte Carlo
Synthetic-Acceleration (MCSA) method, can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:MCSA}
\end{subequations}
As in the sequential method, the hat on $\vA$ in
Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the Monte Carlo solution
only approximately inverts this operator.  Thus, we have defined a scheme in
which the initial estimate of $\vx$ in each iteration is updated using a
single fixed-point iteration\footnote{A subspace method could be used instead
  of a stationary method here.}.  The residual is calculated and used as a
source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
error is then used to calculate an updated iterate of the solution vector,
$\vx^{l+1}$.  The entire sequence is iterated to convergence based on the
following stopping criterion~\cite{kelley_1995}:
\begin{equation}
  \|\vr\|_{2} < \epsilon\cdot\|\vb\|_2\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

Before completing the derivation of discrete equations required to solve
Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary conditions.  For
this work, the Marshak boundary condition suffices:
\begin{equation}
  f_b^\pm = \frac{1}{4}\phi -
  \frac{1}{2}\ve{F}\cdot\hat{\ve{n}}\:,\quad \vr\in \partial\Gamma\:,
\end{equation}
where $f_b$ is the incoming partial current on the problem boundary,
$\partial\Gamma$.  On any low boundary we can write the flux using a
first-order discrete form of Fick's Law:
\begin{equation}
  F_{1/2} = -\frac{2}{3\sigma_1}\frac{\phi_1 -
    \phi_{1/2}}{\Delta_1}\:.
  \label{eq:low-boundary-flux}
\end{equation}
As before, we enforce continuity of flux on the boundary using the
Marshak Boundary condition:
\begin{equation}
  -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1} = 2f_b^{+}
  -\frac{1}{2}\phi_{1/2}\:.
\end{equation}
Solving for $\phi_{1/2}$ and applying to Eq.~(\ref{eq:low-boundary-flux})
yields an expression for the low-side boundary flux:
\begin{equation}
  F_{1/2} = \frac{2}{3\sigma_1\Delta_1 + 4}(4f_b^{+}-\phi_1)\:.
\end{equation}
Following an identical procedure on high-side problem boundaries gives
\begin{equation}
  F_{N_l+1/2} = \frac{2}{3\sigma_{N_l}\Delta_{N_l} +
    4}(\phi_{N_l}-4f_b^{-})\:,
\end{equation}
where $l\in\{i,j,k\}$ is defined over the range $[1,N_l]$.

@@

The adjoint Monte Carlo method described in \S~\ref{sec:adjoint-method} is
used to estimate
\begin{equation}
  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
\end{equation}
As noted above, the residual acts as the source for this simulation, and the
operator $\ve{T}$ is implicitly defined in
Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
interpretations that can be applied to this system.  The first follows the
mathematical presentation given in \S~\ref{sec:monte-carlo-matrix}.  Namely,
we form probabilities from the iteration matrix and perform random walks to
generate unbiased estimates of the solution vector, $\delta\bphi^{l+1/2}$,
using the estimator in Eq.~(\ref{eq:adjoint-tally}).

@@

The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
Figure~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
collision results in a scatter to one of the six adjacent cells since
$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
probabilities for each scattering event from
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
the iteration matrix.  In this mesh cell, the probabilities for leakage out of
each face are derived from the equations from each adjacent cell.  Using
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
equations from each adjacent cell, here only noting the contribution from cell
7:
\begin{align*}
  \phi_1 &= \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}
  \frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\
  %%
  \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
  {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
  %%
  \phi_3 &= \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}
  \frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\
  %%
  \phi_4 &= \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}
  \frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
  %%
  \phi_5 &= \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}
  \frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
  %%
  \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:.\\
\end{align*}
A quick view of these equations reveals that the only non-zero entries in
column 7 of the iteration matrix are
\begin{equation*}
  \begin{pmatrix}
    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
  \end{pmatrix}^{T}
  =
  \begin{pmatrix}
    \frac{\sigm_{7\rightarrow
        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
    \\ \frac{\sigp_{7\rightarrow
        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
    \\ \frac{\sigm_{7\rightarrow
        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
    \\ \frac{\sigp_{7\rightarrow
        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
    \\ \frac{\sigm_{7\rightarrow
        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
    \\ \frac{\sigp_{7\rightarrow
        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
  \end{pmatrix}\:,
\end{equation*}
where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI - \ve{T})$ is
the iteration matrix.  Now, using Eq.~(\ref{eq:adjoint-probability}), we
define the scattering probabilities for cell 7 as
\begin{equation*}
  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
  \{1,\ldots,6\}\:.
\end{equation*}
The weight change for each scattering event is calculated using
Eq.~(\ref{eq:adjoint-weight}):
\begin{equation*}
  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
  \{1,\ldots,6\}\:,
\end{equation*}
where $m$ is the step-index in the particle history.  We also note that, for
the 3D Cartesian mesh considered in this work, the maximum number of non-zero
entries in columns of the iteration matrix is 6, generating a sparse system.
Cells along the problem boundary can have fewer than six entries.
\begin{figure}[ht!]
  \centerline{ \includegraphics[clip]{3D_mesh_segment.pdf}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}
